E-Graphs With Bindings

### 2025-01-26 LICS_2025/e_graphs_with_bindings/sections/combinatorial_semantics.tex
made formatting more compact by inlining some maths

In this section,  we recall the fundamental definitions and results of string diagram rewrite theory for symmetric monoidal categories,  recapitulating the correspondence between string diagram rewriting and an appropriate notion of double pushout (DPO) rewriting of certain hypergraphs,  as established in~\cite{bonchi_string_2022-1, bonchi_string_2022-2}.

@@

\begin{definition}[Category of hypergraphs]\label{def:hypergraph}
A \emph{hypergraph $\mathcal{G}$ over monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$} is a tuple $(V,E,s,t,l)$,  where $V$ is a finite set of vertices, $E$ is a finite set of edges, $s : E \to V^{*}$ is a source function, $t : E \to V^{*}$ is a target function,  and $l = (l_{V} : V \to \Sigma_{O},l_{E} : E \to \Sigma_{M})$ is a pair of labelling functions that assigns each edge and vertex an operator and a type from monoidal signature $\Sigma$ respectively.  
The labelling function must respect the typing of the operator: for all edges $e$, such that $l(e) = c : A \to B$,  $l_{V}^{*}(s(e)) = A$ and $l_{V}^{*}(t(e)) = B$.  
We call a hypergraph \textit{discrete} if its set of edges is empty.   
A \emph{hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ is given by a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that the following hold. 
\begin{enumerate}
    \item $\phi_V^*(s_{\mathcal{F}}(e)) = s_{\mathcal{G}}(\phi_E(e)) \qquad \phi_V^*(t_{\mathcal{F}}(e)) = t_{\mathcal{G}}(\phi_E(e))$
    \item $l_{V,\mathcal{F}}(v) = l_{V,\mathcal{G}}(\phi_V(v)) \qquad\;\; l_{E,\mathcal{F}}(e) = l_{E,\mathcal{G}}(\phi_E(e))$
\end{enumerate}
That is, a homomorphism should preserve sources and targets of edges and preserve the labels.
We denote by $\catname{Hyp(\Sigma)}$ the category of hypergraphs over $\Sigma$ and hypergraph homomorphisms.
\end{definition}
$\catname{Hyp(\Sigma)}$ has all finite colimits and, in particular,  the coproduct is given by the disjoint union of hypergraphs.
The initial object is given by the empty hypergraph.

@@

Composition of two cospans $n \xrightarrow{f} \mathcal{F} \xleftarrow{g} m;m \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} k$ is given by the pushout
\[
n \xrightarrow{f} \mathcal{F} +_{g,f'} \mathcal{G} \xleftarrow{g'} k
\]
that is computed by identifying vertices in $\mathcal{F} + \mathcal{G}$ that share pre-image in $m$.
As pushout is only defined up to isomorphism, it justifies the morphisms in this category being isomorphism classes of cospans.

To make such cospans complete with respect to representing morphism of $\catname{S}(\Sigma)$, we need to impose some restriction on their form.
\begin{definition}[Degree,  path,  acyclic] 
The \emph{in-degree} of a vertex $v$ in a hypergraph $\mathcal{G}$ is the number of hyperedges $e \in {E_\mathcal{G}}$ such that $v \in t(e)$.  
Similarly, the out-degree of $v$ is the number of edges $e \in E_\mathcal{G}$ such that $v \in s(e)$.
A \emph{path} from $e_0$ to $e_{n-1}$ of length $n$ in a hypergraph $\mathcal{G}$ is an ordered sequence of edges $[e_0, e_1, \ldots, e_{n-1}] \in E_\mathcal{G}^*$ such that for all $i < n - 1$ some vertex $v \in t(e_i)$ is also in $s(e_{i+1})$.
A hypergraph is \emph{directed acyclic} if it contains no paths of length $n > 0$ that start and end in the same edge.
\end{definition}

\begin{definition}[Category of MDA Hypergraphs with Interfaces~\cite{bonchi_string_2022-2}]
\label{def:monogamy_hyp}
We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\HypI{\Sigma}$ \emph{monogamous directed acyclic (MDA)} if:
\begin{enumerate}
    \item $\mathcal{G}$ is a directed acyclic hypergraph;
    \item $f$ and $g$ are monomorphisms;
    \item the in-degree and out-degree of every vertex is at most $1$;
    \item vertices with in-degree $0$ are precisely the image of $f$; and
    \item vertices with out-degree $0$ are precisely the image of $g$.
\end{enumerate}
We denote by $\MdaCospans$ the subcategory of $\HypI{\Sigma}$ of MDA cospans.
\end{definition}
Note that $\MdaCospans$ remains a symmetric monoidal category with the same equipment as before.  We now recall from
\cite{bonchi_string_2022-2} the following theorem,  which says the category $\MdaCospans$ is in fact \textit{equivalent} to $\catname{S}(\Sigma)$. 
\begin{theorem}[Corollary 26 of~\cite{bonchi_string_2022-2}]\label{thm:prop-equiv}
    $\catname{S}(\Sigma) \cong \MdaCospans$
\end{theorem}

@@

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For MDA cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} \mathcal G \xleftarrow{b_2} m$
and mono $f : \mathcal L \to \mathcal G$, a pushout complement $i + j \to \mathcal{L}^\bot \to \mathcal G$ as depicted in the square below
\[
    \adjustbox{scale=0.8}{
        \tikzfig{./figures/DPOI_boundary_complement}
    }\]
is called a boundary complement if $[c_1, c_2]$ is mono and there exist $d_1 : n \to \mathcal{L}^{\bot}$ and $d_2 : m \to \mathcal{L}^{\bot}$ making the above triangle commute and such that
\[
    n + j \xrightarrow{[d_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[d_2,c_1]} m + i
\]
is an MDA cospan. 
\end{definition}

@@

\begin{definition}[Convex match]
We call a subgraph $\mathcal{H}$ of a hypergraph $\mathcal{G}$ a \emph{convex subgraph} if for all $v_i, v_j \in V_{\mathcal{H}}$ every path from $v_i$ to $v_j$ is also in $\mathcal{H}$.
We call a match $f : \mathcal L \to \mathcal G$ a \emph{convex match} if it is mono and its image is a convex subgraph of $\mathcal G$.

@@

\begin{definition}
We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{red}{<}, \textcolor{blue}{<},\consistency)$ where
\begin{itemize}
  \item $V$ is a set of vertices.
  \item $E = \colorbox{yellow}{E} \cup \textcolor{red}{E} \cup \textcolor{blue}{E}$ is a set of hyperedges that that is formed of three disjoint sets of hyperedges defined below.
  \item $s,t : E \to V^{*}$ are source and target functions.
  \item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \Sigma_{O} \cup \text{obj}_{\multimap}$ are \textit{label} functions, where $\text{obj}_{\multimap}$ consists of objects $A \multimap B$ for $A,B \in obj_{\Sigma_{O}}$.
  \item $\textcolor{red}{<} \subseteq \textcolor{red}{E} \times V + E$ and $\textcolor{blue}{<} \subseteq \textcolor{blue}{E} \times V + E $ with $\textcolor{red}{E}, \textcolor{blue} \subset E$ are disjoint partial orders that are defined as a transitive closure of corresponding immediate predecessor relations that we denote as $\textcolor{red}{<^{\mu}}$ and $\textcolor{blue}{<^{\mu}}$.
        For $x \in V + E$, the set of predecessors of $x$ is denoted as $[x) = \{x' ~|~ \exists y_{1} = x', \ldots, y_{n} = x,  y_{i} \textcolor{red}{<^{\mu}} y_{i + 1} \text{ or } y_{i} \textcolor{blue}{<^{\mu}} y_{i+1}\; \}$.
        We call edges $e$ such that $l(e) = \bot$ hierarchical and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \textit{top-level}.
        Each of these relations should satisfy the following
        \begin{enumerate}
          \item each set of predecessors consists exclusively of hierarchical edges;
          \item each $x$ has at most one immediate predecessor;
          \item edges $e$ such that there is no $x$ such that $e \textcolor{red}{<^{\mu}} x$ or $e \textcolor{red}{<^{\mu}} x$ are labelled;
          \item the relations are closed under connectivity, \emph{i.e.}, if $v \in s(e)$ then $e' \textcolor{red}{<^{\mu}} e$ iff $e' \textcolor{red}{<^{\mu}} v$, similarly for $v \in t(e)$ and $\textcolor{blue}{<}$.
        \end{enumerate}
  \item $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p \textcolor{red}{<^\mu} x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p \textcolor{red}{<^{\mu}}(v)$ and $p \textcolor{red}{<^{\mu}}(e)$ then $v \consistency_{p} e$.
  We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p \textcolor{red}{<^{\mu}} v\}$ and $E_{p} = \{ e ~ | ~ p \textcolor{red}{<^{\mu}} e\}$.
\end{itemize} 
$\textcolor{red}{E}$ and $\textcolor{blue}{E}$ are then induced by the corresponding relations and consist of exclusively hierarchical edges; then we have $\colorbox{yellow}{E} = E \setminus \textcolor{red}{E} \cup \textcolor{blue}{E}$.
Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $\textcolor{red}{<}, \textcolor{blue}{<}$ and~$\consistency$.
\end{definition}

Intuitively, edges from $\textcolor{red}{E}$ encode equivalence classes, \emph{i.e.}, e-boxes, while edges from $\textcolor{blue}{E}$ encode lambda-abstraction boxes.
Also note the typing on the $l_{V}$ function.
We work with freely generated closed semilattice-enriched categories, and to avoid unnecessary complications we limit the labelling of vertices to such a type.
In particular, it prohibits vertices to be labelled with, \emph{e.g.}, $A \otimes B$, but allows them to be labelled with $A \otimes B \multimap C \otimes D$.
$\textcolor{blue}{<}$ defines closed monoidal structure as defined in~\cite{fscd} and $(\textcolor{red}{<},\consistency)$ defines $\catname{SLat}$ structure as per~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.

Consider an example in the middle part of Figure~\ref{fig:e-cospan-example}.
The e-hypergraph in the example is given by $V = \{v_{1}, \ldots w_{10}\}$, $E = \{e_{1}, \ldots, e_{7}\}$, $\textcolor{red}{E} = \{e_{1}\}$, $\textcolor{blue}{E} = \{e_{2}\}$, $l_{E} = {e_{1} \mapsto \bot, e_{2} \mapsto \bot, e_{3} \mapsto g, \ldots, e_{7} \mapsto f}$ and then we have
\[
\vspace{-1ex}
\begin{array}{cccc}
  v_{4} \textcolor{red}{<^{\mu}} e_{1} & v_{7} \textcolor{blue}{<^{\mu}} e_{2} & v_{4} \consistency e_{2} & v_{4} \not \consistency v_{10}\\
  e_{3} \textcolor{red}{<^{\mu}} e_{1} & v_{8} \textcolor{blue}{<^{\mu}} e_{2} &  e_{2} \consistency u_{1} & e_{2} \not \consistency e_{6}\\
  e_{2} \textcolor{red}{<^{\mu}} e_{1} & e_{5} \textcolor{blue}{<^{\mu}} e_{2} & e_{2} \consistency e_{4} & e_{4} \not \consistency e_{7}\\  
  \ldots & \ldots & \ldots & \ldots \\
  v_{12} \textcolor{red}{<^{\mu}} e_{1} & w_{9} \textcolor{blue}{<^{\mu}} e_{2} & v_{11} \consistency v_{12} & w_{6} \not \consistency w_{10}
\end{array}
\]
and, in particular, we have $[e_{5}) = \{e_2, e_1\}$.
$\consistency_{e_{1}}$ defines a partition on immediate successors of $e_{1}$ which we delimit by a vertical dashed line.
Edges in $\textcolor{red}{E}$ are depicted with a dashed box, while edges in $\textcolor{blue}{E}$ --- with a round solid box.
\begin{figure}
\[
\adjustbox{scale=0.6}{
\tikzfig{./figures/closed_iso_example_2}
}
\]
\captionsetup{skip=0pt, belowskip=-1ex}
\caption{Cospan of e-hypergraphs example}
\label{fig:e-cospan-example}
\end{figure}
\begin{definition}[E-hypergraph homomorphism]
  An \textit{e-hypergraph homomorphism} $\phi:\mathcal{F}\to\mathcal{G}$ between e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
  \begin{enumerate}
      \item $\phi$ is a hypergraph homomorphism;
      \item $\phi_{E}(\textcolor{red}{E_{\mathcal{F}}}) \subseteq \textcolor{red}{E_{\mathcal{G}}}$ and $\phi_{E}(\textcolor{blue}{E_{\mathcal{F}}}) \subseteq \textcolor{blue}{E_{\mathcal{G}}}$;
      \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
        \begin{align*}
          \text{if } e \textcolor{red}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{red}{<_{\mathcal{G}}^{\mu}} \phi(v)
          \qquad
          \text{if } e \textcolor{blue}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{blue}{<_{\mathcal{G}}^{\mu}} \phi(v)
        \end{align*}
          and for $e_1, e_2 \in E_{\mathcal{F}}$
          \begin{align*}
          \text{if } e_1 \textcolor{red}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{red}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \qquad
          \text{if } e_1 \textcolor{blue}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{blue}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \end{align*}
          \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$.
  \end{enumerate}
  % If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
  \end{definition}
  The conditions on e-hypergraph homomorphisms require preserving immediate predecessors and the consistency relation.
  \begin{definition}[Category of e-hypergraphs]
  The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
  \end{definition}

The category of e-hypergraphs has coproducts,  given by the disjoint union of e-hypergraphs,  and an initial object given by the empty e-hypergraph.  
A concrete description of the pushout of two morphisms in this category is given in Appendix~\ref{sec:appendix:pushout}.
Generally,  the pushout of two e-hypergraph homomorphisms need not exist,  but it does when certain technical conditions are satisfied. 
Most importantly a pushout for the composition of two cospans with discrete feet always exists, as described below.

@@

\begin{definition}[Isomorphic cospans]
Consider a relation 
\begin{multline*}
R = \{ x R y \text{ if } f_{\text{int}}(x) \consistency f_{\text{int}}(y)\\ \;\text{or}\; \exists z \;\text{s.t.}\; z \textcolor{blue}{<^{\mu}} f_{\text{int}}(x) \text{ and } z \textcolor{blue}{<}^{\mu} f_{\text{int}}(y)\}
\end{multline*}
for $x, y \in n'$.
And let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}^{k}_{j=1}$.
We get an analogous partition for $m'$.

@@

For MDA cospans 
\[
    i \xrightarrow{} i' \xrightarrow{} \mathcal{L} \xleftarrow{} j' \xleftarrow{} j
\quad\text{ and }\quad
    n \xrightarrow{} n' \xrightarrow{} \mathcal{G} \xleftarrow{} k' \xleftarrow{} k
\] and mono $m : \mathcal{L} \to \mathcal{G}$ in $\catname{EHyp(\Sigma)}$, a pushout complement $i + j \to \mathcal{L}^{\bot} \to \mathcal{G}$
as depicted in the square below
\[
\scalebox{0.75}{

>>>>

We recall the fundamental definitions and results of string diagram rewrite theory for symmetric monoidal categories,  recapitulating the correspondence between string diagram rewriting and an appropriate notion of double pushout (DPO) rewriting of certain hypergraphs,  as established in~\cite{bonchi_string_2022-1, bonchi_string_2022-2}.

@@

\begin{definition}[Category of hypergraphs]\label{def:hypergraph}
A \emph{hypergraph $\mathcal{G}$ over monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$} is a tuple $(V,E,s,t,l)$,  where $V$ is a finite set of vertices, $E$ is a finite set of edges, $s : E \to V^{*}$ is a source function, $t : E \to V^{*}$ is a target function,  and $l = (l_{V} : V \to \Sigma_{O},l_{E} : E \to \Sigma_{M})$ is a pair of labelling functions that assigns each edge and vertex an operator and a type from monoidal signature $\Sigma$ respectively.  
The labelling function must respect the typing of the operator: for all edges $e$, such that $l(e) = c : A \to B$,  $l_{V}^{*}(s(e)) = A$ and $l_{V}^{*}(t(e)) = B$.  
We call a hypergraph \textit{discrete} if its set of edges is empty.   
A \emph{hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ is given by a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that the following hold. 
\begin{enumerate}
    \item $\phi_V^*(s_{\mathcal{F}}(e)) = s_{\mathcal{G}}(\phi_E(e)) \qquad \phi_V^*(t_{\mathcal{F}}(e)) = t_{\mathcal{G}}(\phi_E(e))$
    \item $l_{V,\mathcal{F}}(v) = l_{V,\mathcal{G}}(\phi_V(v)) \qquad\;\; l_{E,\mathcal{F}}(e) = l_{E,\mathcal{G}}(\phi_E(e))$
\end{enumerate}
\end{definition}
We denote by $\catname{Hyp(\Sigma)}$ the category of hypergraphs over $\Sigma$ and hypergraph homomorphisms.
$\catname{Hyp(\Sigma)}$ has all finite colimits and, in particular,  the coproduct is given by the disjoint union of hypergraphs.
The initial object is given by the empty hypergraph.

@@

Composition of two cospans $n \xrightarrow{f} \mathcal{F} \xleftarrow{g} m;m \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} k$ is given by the pushout
$
n \xrightarrow{f} \mathcal{F} +_{g,f'} \mathcal{G} \xleftarrow{g'} k
$,
computed by identifying vertices in $\mathcal{F} + \mathcal{G}$ that share pre-image in $m$.
As pushout is only defined up to isomorphism, it justifies the morphisms in this category being isomorphism classes of cospans.
To make such cospans complete with respect to representing morphism of $\catname{S}(\Sigma)$, we need to impose some restriction on their form.
\begin{definition}[Degree,  path,  acyclic] 
The \emph{in-degree} of a vertex $v$ in a hypergraph $\mathcal{G}$ is the number of hyperedges $e \in {E_\mathcal{G}}$ such that $v \in t(e)$.  
Similarly, the out-degree of $v$ is the number of edges $e \in E_\mathcal{G}$ such that $v \in s(e)$.
A \emph{path} from $e_0$ to $e_{n-1}$ of length $n$ in a hypergraph $\mathcal{G}$ is an ordered sequence of edges $[e_0, e_1, \ldots, e_{n-1}] \in E_\mathcal{G}^*$ such that for all $i < n - 1$ some vertex $v \in t(e_i)$ is also in $s(e_{i+1})$.
A hypergraph is \emph{directed acyclic} if it contains no paths of length $n > 0$ that start and end in the same edge.
\end{definition}

\begin{definition}[Category of MDA Hypergraphs with Interfaces~\cite{bonchi_string_2022-2}]
\label{def:monogamy_hyp}
We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\HypI{\Sigma}$ \emph{monogamous directed acyclic (MDA)} if:
\begin{enumerate}
    \item $\mathcal{G}$ is a directed acyclic hypergraph;
    \item $f$ and $g$ are monomorphisms;
    \item the in-degree and out-degree of every vertex is at most~$1$;
    \item vertices with in-degree $0$ are precisely the image of $f$; 
    \item vertices with out-degree $0$ are precisely the image of $g$.
\end{enumerate}
We denote by $\MdaCospans$ the subcategory of $\HypI{\Sigma}$ of MDA cospans.
\end{definition}
Note that $\MdaCospans$ remains a symmetric monoidal category with the same equipment as before.  We now recall from
\cite{bonchi_string_2022-2} the following theorem,  which says the category $\MdaCospans$ is in fact \textit{equivalent} to $\catname{S}(\Sigma)$. 
\begin{theorem}[Corollary 26 of~\cite{bonchi_string_2022-2}]\label{thm:prop-equiv}
    $\catname{S}(\Sigma) \cong \MdaCospans$
\end{theorem}

@@

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For MDA cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} \mathcal G \xleftarrow{b_2} m$
and mono $f : \mathcal L \to \mathcal G$, a pushout complement $i + j \to \mathcal{L}^\bot \to \mathcal G$ as depicted in the square below
\[
    \adjustbox{scale=0.8}{
        \tikzfig{./figures/DPOI_boundary_complement}
    }\]
is called a boundary complement if $[c_1, c_2]$ is mono and there exist $d_1 : n \to \mathcal{L}^{\bot}$ and $d_2 : m \to \mathcal{L}^{\bot}$ making the above triangle commute and such that
$
    n + j \xrightarrow{[d_1,c_2]} \mathcal{L}^{\bot} \xleftarrow{[d_2,c_1]} m + i
$
is an MDA cospan. 
\end{definition}

@@

\begin{definition}[Convex match]
A subgraph $\mathcal{H}$ of a hypergraph $\mathcal{G}$ is a \emph{convex subgraph} if for all $v_i, v_j \in V_{\mathcal{H}}$ every path from $v_i$ to $v_j$ is also in $\mathcal{H}$.
A match $f : \mathcal L \to \mathcal G$ is a \emph{convex match} if it is mono and its image is a convex subgraph of $\mathcal G$.

@@

\begin{definition}
We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{orange}{<}, \textcolor{blue}{<},\consistency)$ where
\begin{itemize}
  \item $V$ is a set of vertices.
  \item $E = \colorbox{yellow}{E} \cup \textcolor{orange}{E} \cup \textcolor{blue}{E}$ is a set of hyperedges that that is formed of three disjoint sets of hyperedges defined below.
  \item $s,t : E \to V^{*}$ are source and target functions.
  \item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \Sigma_{O} \cup \text{obj}_{\multimap}$ are \textit{label} functions, where $\text{obj}_{\multimap}$ consists of objects $A \multimap B$ for $A,B \in obj_{\Sigma_{O}}$.
  \item $\textcolor{orange}{<} \subseteq \textcolor{orange}{E} \times V + E$ and $\textcolor{blue}{<} \subseteq \textcolor{blue}{E} \times V + E $ with $\textcolor{orange}{E}, \textcolor{blue} \subset E$ are disjoint partial orders that are defined as a transitive closure of corresponding immediate predecessor relations that we denote as $\textcolor{orange}{<^{\mu}}$ and $\textcolor{blue}{<^{\mu}}$.
        For $x \in V + E$, the set of predecessors of $x$ is denoted as $[x) = \{x' ~|~ \exists y_{1} = x', \ldots, y_{n} = x,  y_{i} \textcolor{orange}{<^{\mu}} y_{i + 1} \text{ or } y_{i} \textcolor{blue}{<^{\mu}} y_{i+1}\; \}$.
        We call edges $e$ such that $l(e) = \bot$ hierarchical and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \textit{top-level}.
        Each of these relations should satisfy the following
        \begin{enumerate}
          \item each set of predecessors consists exclusively of hierarchical edges;
          \item each $x$ has at most one immediate predecessor;
          \item edges $e$ such that there is no $x$ such that $e \textcolor{orange}{<^{\mu}} x$ or $e \textcolor{orange}{<^{\mu}} x$ are labelled;
          \item the relations are closed under connectivity, \emph{i.e.}, if $v \in s(e)$ then $e' \textcolor{orange}{<^{\mu}} e$ iff $e' \textcolor{orange}{<^{\mu}} v$, similarly for $v \in t(e)$ and $\textcolor{blue}{<}$.
        \end{enumerate}
  \item $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p \textcolor{orange}{<^\mu} x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p \textcolor{orange}{<^{\mu}}(v)$ and $p \textcolor{orange}{<^{\mu}}(e)$ then $v \consistency_{p} e$.
  We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p \textcolor{orange}{<^{\mu}} v\}$ and $E_{p} = \{ e ~ | ~ p \textcolor{orange}{<^{\mu}} e\}$.
\end{itemize} 
$\textcolor{orange}{E}$ and $\textcolor{blue}{E}$ are then induced by the corresponding relations and consist of exclusively hierarchical edges, so $\colorbox{yellow}{E} = E \setminus \textcolor{orange}{E} \cup \textcolor{blue}{E}$.
Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $\textcolor{orange}{<}, \textcolor{blue}{<}$ and~$\consistency$.
\end{definition}

Intuitively, edges from $\textcolor{orange}{E}$ encode equivalence classes, \emph{i.e.}, e-boxes, while edges from $\textcolor{blue}{E}$ encode lambda-abstraction boxes.
Also note the typing on the $l_{V}$ function.
We work with freely generated closed semilattice-enriched categories, and to avoid unnecessary complications we limit the labelling of vertices to such a type.
In particular, it prohibits vertices to be labelled with, \emph{e.g.}, $A \otimes B$, but allows them to be labelled with $A \otimes B \multimap C \otimes D$.
$\textcolor{blue}{<}$ defines closed monoidal structure as defined in~\cite{fscd} and $(\textcolor{orange}{<},\consistency)$ defines $\catname{SLat}$ structure as per~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.

Consider an example in the middle part of Figure~\ref{fig:e-cospan-example}.
The e-hypergraph in the example is given by $V = \{v_{1}, \ldots w_{10}\}$, $E = \{e_{1}, \ldots, e_{7}\}$, $\textcolor{orange}{E} = \{e_{1}\}$, $\textcolor{blue}{E} = \{e_{2}\}$, $l_{E} = {e_{1} \mapsto \bot, e_{2} \mapsto \bot, e_{3} \mapsto g, \ldots, e_{7} \mapsto f}$ and then we have
\[
\vspace{-1ex}
\begin{array}{cccc}
  v_{4} \textcolor{orange}{<^{\mu}} e_{1} & v_{7} \textcolor{blue}{<^{\mu}} e_{2} & v_{4} \consistency e_{2} & v_{4} \not \consistency v_{10}\\
  e_{3} \textcolor{orange}{<^{\mu}} e_{1} & v_{8} \textcolor{blue}{<^{\mu}} e_{2} &  e_{2} \consistency u_{1} & e_{2} \not \consistency e_{6}\\
  e_{2} \textcolor{orange}{<^{\mu}} e_{1} & e_{5} \textcolor{blue}{<^{\mu}} e_{2} & e_{2} \consistency e_{4} & e_{4} \not \consistency e_{7}\\  
  \ldots & \ldots & \ldots & \ldots \\
  v_{12} \textcolor{orange}{<^{\mu}} e_{1} & w_{9} \textcolor{blue}{<^{\mu}} e_{2} & v_{11} \consistency v_{12} & w_{6} \not \consistency w_{10}
\end{array}
\]
and, in particular, we have $[e_{5}) = \{e_2, e_1\}$.
$\consistency_{e_{1}}$ defines a partition on immediate successors of $e_{1}$ which we delimit by a vertical dashed line.
Edges in $\textcolor{orange}{E}$ are depicted with a dashed box, while edges in $\textcolor{blue}{E}$ --- with a round solid box.
\begin{figure}
\[
\adjustbox{scale=0.6}{
\tikzfig{./figures/closed_iso_example_2}
}
\]
\captionsetup{skip=0pt, belowskip=-1ex}
\caption{Cospan of e-hypergraphs example}
\label{fig:e-cospan-example}
\end{figure}
\begin{definition}[E-hypergraph homomorphism]
  An \textit{e-hypergraph homomorphism} $\phi:\mathcal{F}\to\mathcal{G}$ between e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
  \begin{enumerate}
      \item $\phi$ is a hypergraph homomorphism;
      \item $\phi_{E}(\textcolor{orange}{E_{\mathcal{F}}}) \subseteq \textcolor{orange}{E_{\mathcal{G}}}$ and $\phi_{E}(\textcolor{blue}{E_{\mathcal{F}}}) \subseteq \textcolor{blue}{E_{\mathcal{G}}}$;
      \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
        \begin{align*}
          \text{if } e \textcolor{orange}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{orange}{<_{\mathcal{G}}^{\mu}} \phi(v)
          \qquad
          \text{if } e \textcolor{blue}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{blue}{<_{\mathcal{G}}^{\mu}} \phi(v)
        \end{align*}
          and for $e_1, e_2 \in E_{\mathcal{F}}$
          \begin{align*}
          \text{if } e_1 \textcolor{orange}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{orange}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \qquad
          \text{if } e_1 \textcolor{blue}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{blue}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \end{align*}
          \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$.
  \end{enumerate}
  % If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
  \end{definition}
  The conditions on e-hypergraph homomorphisms require preserving immediate predecessors and the consistency relation.
  \begin{definition}[Category of e-hypergraphs]
  The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
  \end{definition}

The category of e-hypergraphs has coproducts,  given by the disjoint union of e-hypergraphs,  and an initial object given by the empty e-hypergraph.  
A concrete description of the pushout of two morphisms in this category is given in Appendix~\ref{sec:appendix:pushout}.
Generally,  the pushout of two e-hypergraph homomorphisms need not exist,  but it does when certain technical conditions are satisfied. 
However, a pushout for the composition of two cospans with discrete feet always exists, as described below.

@@

\begin{definition}[Isomorphic cospans]
Consider a relation 
\begin{multline*}
R = \bigl\{ x R y \text{ if } f_{\text{int}}(x) \consistency f_{\text{int}}(y)\\ \;\text{or}\; \exists z \;\text{s.t.}\; z \textcolor{blue}{<^{\mu}} f_{\text{int}}(x) \text{ and } z \textcolor{blue}{<}^{\mu} f_{\text{int}}(y)\bigr\}
\end{multline*}
for $x, y \in n'$.
And let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}^{k}_{j=1}$.
We get an analogous partition for $m'$.

@@

For MDA cospans 
$
    i \xrightarrow{} i' \xrightarrow{} \mathcal{L} \xleftarrow{} j' \xleftarrow{} j
$ and $
    n \xrightarrow{} n' \xrightarrow{} \mathcal{G} \xleftarrow{} k' \xleftarrow{} k
$ and mono $m : \mathcal{L} \to \mathcal{G}$ in $\catname{EHyp(\Sigma)}$, a pushout complement $i + j \to \mathcal{L}^{\bot} \to \mathcal{G}$
as depicted in the square below
\[
\scalebox{0.75}{
